---
title: "determine_upstream_downstream"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
#| output: false 
library(sfnetworks)
library(sf)
library(tidygraph)
library(dplyr)
library(purrr)
library(TSP)
library(tmap)
library(igraph)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(renv)
library(sfnetworks)
```

Load data
```{r}
nepal_rivers <- readRDS("../../../../../capstone/netzerohydro/data/cleaned_v1/nepal_rivers_3up.rds") 
nepal_current_dams <- readRDS("../../../../../capstone/netzerohydro/data/cleaned_v1/nepal_current_dams.rds") 
nepal_future_dams <- readRDS("../../../../../capstone/netzerohydro/data/cleaned_v1/nepal_future_dams.rds") 
```

Change `nepal_rivers` to 'LINESTRING'
```{r}
# Change MULTISTRINGS TO LINESTRING (for sfnetwork)
nepal_rivers <- st_cast(nepal_rivers , "LINESTRING")

plot(nepal_rivers)
```

## Convert Nepal River network to sfnetwork

```{r}
net <- as_sfnetwork(nepal_rivers, direction = TRUE) %>% # convert river geom into spatial network
  tidygraph::activate("edges") %>%  # "Next operation applied to edges table (not node table)"
  mutate(weight = edge_length()) # calculate spatial length of each edge (river segment)

# Look at net  
plot(net, col = "blue", main = "Base R Plot of net from nepal_rivers")
```

Checking direction

```{r}
# Checking if directional
st_network_paths(net, from = 3, to = 15, weight = "weight") 
st_network_paths(net, from = 15, to = 3, weight = "weight") # warned! 

```

Checking CRS and geometry names

```{r}
# Check CRS 
#st_crs(nepal_current_dams) # EPSG",4326
#st_crs(nepal_future_dams) # EPSG",4326
#st_crs(nepal_rivers) # "EPSG",4326 
#st_crs(net) # WGS 84 / EPSG",4326

st_crs(nepal_current_dams) == st_crs(nepal_rivers) # TRUE 
st_crs(nepal_current_dams) == st_crs(net) # TRUE 
st_crs(nepal_future_dams) == st_crs(nepal_rivers) # TRUE 
st_crs(nepal_future_dams) == st_crs(net) # TRUE 

# Change shape geometry to `geometry` 
nepal_rivers <- st_set_geometry(nepal_rivers, "geometry") 

# `nepal_current_dams` is polygons. Need to be points 
nepal_current_dams <- st_point_on_surface(nepal_current_dams)

# before blending, mark which nodes are dams 
nepal_current_dams <- nepal_current_dams %>% 
  mutate(is_dam = TRUE,
         type = "current")

nepal_future_dams <- nepal_future_dams %>% 
  mutate(is_dam = TRUE,
         type = "future")

# join dam datasets together
join1 <- nepal_current_dams %>% st_join(nepal_future_dams, suffix = c("", ".y"))
join2 <- nepal_future_dams %>% st_join(nepal_current_dams, suffix = c("", ".y"))
nepal_dams_all <- join1 %>% 
  bind_rows(join2) %>% 
  select(-c("is_dam.y", "type.y", "x16", "x17", "x18", "x19", "x20"))

colnames(nepal_dams_all)
```

# Snapping dams and creating nodes

`st_network_blend()` does the heavy lifting: it finds the nearest edge for each dam point, snaps the point to it, and splits that edge into two new edges with a new node at the dam location.

-   Blends a point into a network: process of (1) snapping given point in nearest edge in network (2) splitting edge at location (3) adding snapped point as node to network

-   If location of snapped point is already a node, the attributes of point will be joined to the pre-existing node
```{r}
# snapping and creating nodes of dams 
net_with_dams <- net %>% 
  st_network_blend(nepal_dams_all, tolerance = 1000)

net_with_dams_df <- as_tibble(net_with_dams) # makes sfnetwork a df

# Plotting every line edge and junction from net_with_dams
ggplot() +
  geom_sf(data = st_as_sf(net_with_dams, "nodes"), color = "red", size = 0.3) +
  geom_sf(data = st_as_sf(net_with_dams, "edges"), color = "gray40", size = 0.3) +
  theme_minimal() +
  labs(title = "net_with_dams with ggplot2")
```

### Look at snapped dams that are now nodes

```{r}

# Which nodes are dams and making them a df 
dam_nodes_sf <- net_with_dams %>% 
  tidygraph::activate("nodes") %>% 
  as_tibble() %>% 
  filter(is_dam == TRUE) %>% 
  st_as_sf()

# Snapped 412 dam nodes (of the 497 `nepal_dams_all` obs) - thats good! 

# Pulling `hybas_id` dams that snapped to the river network 
grand_id_snapped_dams <- net_with_dams %>% 
  tidygraph::activate("nodes") %>% 
  as_tibble() %>% 
  filter(is_dam == TRUE) %>% 
  pull(grand_id)


# Plotting every line edge and juntion from net_with_dams
ggplot() +
  geom_sf(data = st_as_sf(net_with_dams, "edges"), color = "gray40", size = 0.2) +
  geom_sf(data = st_as_sf(dam_nodes_sf, "nodes"), color = "red", size = 0.5) +
  theme_minimal() +
  labs(title = "current and futuredams that became nodes with ggplot2")

```

### Look at total dams numbers vs snapped dams numbers

```{r}
# all dams 
length(nepal_dams_all$grand_id)

# Dams that were snapped 
length(grand_id_snapped_dams)

```

## Creating a **directional** connectivity matrix

```{r}

# Getting the dam node ID 
dam_nodes <- net_with_dams %>% 
  tidygraph::activate("nodes") %>% 
  mutate(node_id = row_number()) %>% # create new column with node ID
  as_tibble() %>% 
  filter(is_dam == TRUE) %>% 
  pull(node_id) # list of all dam nodes 

# Create connective matrix 
connectivity_matrix  <- distances(  # library(igraph) - calculates shortest path 
  net_with_dams, 
  v = dam_nodes, # starting nodes 
  to = dam_nodes, # ending nodes - creats all pairwise combinations 
  weights = tidygraph::activate(net_with_dams, "edges") %>% # use river segment lengths to calculate shortest path 
  pull(weight), # extract the `weight` column
  mode = "out" # DIRECTIONAL CONNECTIVITY 
  )

connectivity_matrix

```

```{r}
# Get hybas_id for each dam node directly from network
dam_grand_ids <- net_with_dams %>%
  tidygraph::activate("nodes") %>%
  as_tibble() %>%
  slice(dam_nodes) %>%
  pull(grand_id)

# Create dam labels using full grand_id
dam_labels <- paste0("Dam_", dam_grand_ids)

# Get number of dams (rows/columns in symmetric matrix)
n_dams <- nrow(connectivity_matrix)

# Create row indices for upper triangle: repeat 1, then 1,1, then 1,1,1, etc.
row_idx <- rep(1:(n_dams-1), times = (n_dams-1):1)

# Create column indices for upper triangle: 2:n, then 3:n, then 4:n, etc.
col_idx <- unlist(lapply(2:n_dams, function(i) i:n_dams))

# Extract distance values from matrix using row/column index pairs
distances <- connectivity_matrix[cbind(row_idx, col_idx)]

# Create dataframe with columns: from dam, to dam, and distance
reach_df <- data.frame(
  from_dam = dam_labels[row_idx],           # Source dam label (hybas_id)
  to_dam = dam_labels[col_idx],             # Destination dam label (hybas_id)
  distance_m = distances,                   # Distance value in meters
  stringsAsFactors = FALSE                  # Keep character columns as character
)

# Count pairs with zero distance (dams at same network node)
zero_pairs <- sum(reach_df$distance_m == 0, na.rm = TRUE)

# Count pairs with duplicate distances (same distance value appears multiple times)
duplicate_distances <- sum(duplicated(reach_df$distance_m[reach_df$distance_m > 0]))

# Filter out zeros (dams at same location) and NA/Inf values (unreachable pairs)
reach_df <- reach_df[reach_df$distance_m > 0 & !is.na(reach_df$distance_m) & is.finite(reach_df$distance_m), ]

# Sort dataframe by distance column (ascending order)
reach_df <- reach_df[order(reach_df$distance_m), ]

# Look at dataframe 
head(reach_df)

# Create histogram plot: x-axis is distance in km
ggplot(reach_df, aes(x = distance_m / 1000)) +
  # Add histogram bars with 30 bins, blue fill
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  # Set axis labels and title
  labs(
    x = "Reach Distance (km)",              # X-axis label (convert m to km)
    y = "Count",                            # Y-axis label (number of pairs)
    title = "Distribution of Reach Distances Between Current Dams"  # Plot title (current dams only)
  ) +
  # Use minimal theme (clean, simple appearance)
  theme_minimal()
```

