---
title: "river_connection_directional"
format: html
---

Load Libraries

```{r}
#| output: false 
library(sfnetworks)
library(sf)
library(tidygraph)
library(dplyr)
library(purrr)
library(TSP)
```

Load Data 
```{r}
trial_river <- readRDS("../../../../../capstone/netzerohydro/data/cleaned_v1/trial_river.rds")


```

```{r}
# Change MULTISTRINGS TO LINESTRING (for sfnetwork)
trial_river <- st_cast(trial_river , "LINESTRING")

```



## Transform data to `sfnetwork` type and calculates edge weights

-   Spatial networks have 2 data tables:

    -   Nodes (junctions/endpoints)
    -   Edges (line connecting points)

-   Create another column: `weight` for river segment *distance*

-   `direction = TRUE` means flow can only do in the direction the line geometeries are drawn (downstream only). Should align with flow direction in HydroRIVERS

    -   df's `next_down` column - each segment points to its downstream neighbor
    -   **For downstream-only paths: This now enforces that paths only follow the direction of line geometries (downstream). A path from upstream→downstream will work, but downstream→upstream won't find a route.**

```{r}
net <- as_sfnetwork(trial_river, direction = TRUE) %>% # convert river geom into spatial network 
  tidygraph::activate("edges") %>%  # "Next operation applied to edges table (not node table)"
  mutate(weight = edge_length()) # calculate spatial length of each edge (river segment)
         

net_df <- as_tibble(net) # makes sfnetwork a df 
```

**Direction:** When you call `net`, it says 'A rooted tree' which means its directed (a tree structure has a root and flows one way).

```{r}
net
```

## Define path

-   Shortest path from node \# (3) to \# (15) using `weight` column

    -   Would count edges \[fewest segments\] instead of distance if not `weight`
    -   \# = Node IDs (row number in nodes table)

-   Returns: list of nodes that are the shortest path/route (EX: `[3, 7, 12, 15]` = path goes thru nodes 3→7→12→15)

-   **Direction** (checking if it worked)

    -   `paths_down` returns a valid path

    -   `paths_up` is invalid and returns a warning!

```{r}
# Define path: find shortest path from node 3 and 15 using weight column
paths_down = st_network_paths(net, from = 3, to = 15, weight = "weight") 

```

```{r}
# Define path: find shortest path from node 15 and 3 (UPSTREAM)
paths_up = st_network_paths(net, from = 15, to = 3, weight = "weight") 
```

## List nodes and edges

```{r}
# lists nodes
paths_down %>%
  slice(1) %>% # first row (first path found)
  pull(node_paths) %>% # extract Node_path (juctions/points you pass thru)
  unlist() # List -> vector 
# lists edges
paths_down %>%
  slice(1) %>%
  pull(edge_paths) %>% # extract edge_path (river segments)
  unlist()

```


## Plot

3 levels to the plotting:

1.  Base layer: plot entire path `plot(net)`

2.  Path: (*use function*)

3.  Start and end node points

```{r}
plot_path = function(node_path) { # input = vector of node IDs as table 
  net %>% 
    activate("nodes") %>% # Node table 
    slice(node_path) %>% # Select nodes in the path 
    plot(cex = 1.5, lwd = 1.5, add = TRUE) # plot thise nodes on top od existing plot 
}


colors = sf.colors(3, categorical = TRUE) # 3 distcict colors for viz 


# Plotting entire network in grey (base layer)
plot(net, col = "grey")

# Plot nodes
paths_down %>%
  pull(node_paths) %>% # extract Node_path (juctions/points you pass thru) of `paths`
  purrr::walk(plot_path) # applies `plot_path` function for each path 

# Plot start and end nodes 
net %>%
  activate("nodes") %>% 
  st_as_sf() %>% # convert nodes to sf object for plotting
  slice(c(3, 15)) %>% # Get only nodes 3 and 15 (start and end)
  plot(col = colors, pch = 8, cex = 2, lwd = 2, add = TRUE) # Plot as stars, in color, on top
```

## Plotting Direction of River

Show flow direction: Each arrow points from line start → midpoint

```{r}
# # Plot base river
# plot(st_geometry(trial_river), col = "grey") # Draws rivee network in grey 
# 
# # Add arrows showing flow direction
# midpoints <- trial_river %>%
#   st_line_sample(sample = 0.5) %>%   # Extract midway point at each line 
#   st_cast("POINT") %>% # point geometery 
#   st_coordinates() %>% # Convert to x,y coorfinate matrix 
#   as.data.frame() 
# 
# # Get start points for arrow direction
# startpoints <- trial_river %>%
#   st_cast("POINT") %>% # covert line -> points 
#   filter(row_number() %% 2 == 1) %>%  # first point of each line (`%% 2 == 1` filters for odd numbered rows, grabbing 1st vertex of each line)
#   st_coordinates() %>% # x,y coordinates 
#   as.data.frame() 
# 
# # Draw arrows
# arrows(startpoints$X, startpoints$Y, # From x and y
#        midpoints$X, midpoints$Y,  # to x and y
#        length = 0.1, col = "blue")

```

